Developing a Bubble Chamber Particle Discriminator Using Semi-Supervised Learning

### 2019-07-25 documentation/paper.tex
Work on paper

Two semi-supervised algorithms were implemented for this study: the existing method pseudo-labels \cite{pseudo}, and a novel method, gravitational differentiation. Both are discussed in detail in Section \ref{semi_supervised}.

@@

The following hyperparameter range was selected based on manual empirical testing, and was thus exhaustively searched:
\begin{itemize}
    \item Dropout $\in \{0, 0.25, 0.5\}$
    \item L2 Regularization $\lambda \in \{0, 0.0003, 0.001, 0.003, 0.01\}$
    \item Number of Dense Layers $\in \{2, 3, 4\}$
    \item Optimizer = Adam \cite{adam}
    \item Loss = Mean Squared Error (found empirically to outperform binary cross-entropy)
    \item Activation = tanh (found empirically to outperform ReLU)
    \item Batch Size = 32
\end{itemize}

@@

Because all data in the PICO-60 experiment is labeled, unlabeled training data sets are randomly sampled from the 496 training examples remaining (after the validation set is selected). Labeled set sizes of 32, 64, 128, and 256 were tested; the remainder of the 496 examples would be used as unlabeled data.

@@

It is a transformation of the hyperbolic tangent that flattens the central range and comparatively exaggerates the asymptote on either side. The equation above can be described in the following steps:

@@

This system was trained using a multi-layer perceptron with an 8-band Fourier transform input for $N(x)$, because this was one of the most successful configurations in supervised learning. It produced the highest accuracy as well as the lowest class-wise standard deviation.

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{grav_diff}
    \caption{\label{grav_diff} A visualization of $\mathrm{GravDiff}(p, \psi, g)$ with respect to $p$ and $\psi$.}
\end{figure}

@@

Table \ref{semi_success_overview} directly compares performance of all semi-supervised learning configurations, with supervised learning, the original PICO-60 ML analysis, and AP provided as baselines for comparison. For every $\varsigma$ size except 32, ICN performs better than supervised learning; for 128 and 256, GD also performs better. For 32, 64, and 128 labeled examples, ICN proves to be more effective than GD; however, when 256 examples are labeled, GD provides the best accuracy, at 99.2\%.

Perhaps most notably, iterative cluster nucleation with only 128 labeled examples achieves accuracy 1.2 percentage points higher than supervised learning with all 496 training examples labeled. It also makes more confident predictions (with a class-wise standard deviation of only 0.21, as opposed to 0.26).

Note that the supervised multi-layer perceptron trained on $\beta _8$ originally used labels for all 496 training examples. Here, it is re-trained with smaller training sets so that it can be more directly compared to semi-supervised learning (which does not have access to labels for the entire set).

\begin{minipage}{\textwidth}
    \begin{center}
        \captionof{table}{Summary of the most successful semi-supervised learning models.} \label{semi_success_overview}
        \begin{tabular}{|l|l|l|l|l|l|}
            \hline
            Initial $\varsigma$ Size & Technique & Accuracy & Precision & Recall & CWSD \\
            \hline
            32 & GD & 71.3\% & 70.1\% & 98.0\% & 0.73 \\
            \hline
            32 & ICN & 85.2\% & 83.4\% & 98.4\% & 0.54 \\
            \hline
            32 & Supervised & 90.4\% & 89.0\% & 97.1\% & 0.53 \\
            \hline
            64 & GD & 79.7\% & 75.5\% & 98.6\% & 0.63 \\
            \hline
            64 & ICN & 94.3\% & 92.2\% & 99.8\% & 0.34 \\
            \hline
            64 & Supervised & 92.7\% & 90.4\% & 98.9\% & 0.41 \\
            \hline
            128 & GD & 97.0\% & 96.3\% & 99.6\% & 0.23 \\
            \hline
            128 & ICN & 98.2\% & 97.7\% & 99.4\% & 0.21 \\
            \hline
            128 & Supervised & 94.7\% & 92.6\% & 99.5\% & 0.37 \\
            \hline
            256 & GD & 99.2\% & 99.0\% & 99.7\% & 0.11 \\
            \hline
            256 & ICN & 98.2\% & 97.6\% & 99.5\% & 0.22 \\
            \hline
            256 & Supervised & 96.5\% & 94.6\% & 99.7\% & 0.28 \\
            \hline
            496 & Supervised & 97.0\% & 95.4\% & 99.9\% & 0.26 \\
            \hline
            278 & PICO-60 ML & 80.7\% & 76.9\% & 100.0\% & 0.65 \\
            \hline
            N/A & AP & 100\% & 100\% & 100\% & 0.43 \\
            \hline
        \end{tabular}
    \end{center}
\end{minipage}
% cat ../experimental_data/training_logs/2018_10_26_banded_ap_similarity_last_300_epochs.out | grep examples256 | grep lambda0.003 | ./mean_ap_accuracy_across_configurations.py
% cat ../experimental_data/training_logs/2018_10_26_banded_ap_similarity_last_300_epochs.out | grep examples128 | grep lambda0.003 | ./mean_ap_accuracy_across_configurations.py
% cat ../experimental_data/training_logs/2018_10_26_banded_ap_similarity_last_300_epochs.out | grep examples64 | grep lambda0.003 | ./mean_ap_accuracy_across_configurations.py
% cat ../experimental_data/training_logs/2018_10_26_banded_ap_similarity_last_300_epochs.out | grep examples32 | grep dropout0.25 | ./mean_ap_accuracy_across_configurations.py

@@

This study has demonstrated that it is possible, using semi-supervised learning, to automatically optimize a discriminator with performance comparable to one developed and tuned by humans like the Acoustic Parameter technique. This conclusion has been achieved through two rounds of optimization: the first to explore and evaluate various network structures and input formats using supervised learning, and the second to incorporate the most successful configuration into semi-supervised learning systems with the goal of producing a high-accuracy discriminator to complement AP.

>>>>

Two semi-supervised algorithms were implemented for this study: the existing method of pseudo-labels \cite{pseudo}, and a novel method, gravitational differentiation. Both are discussed in detail in Section \ref{semi_supervised}.

@@

The following hyperparameter range was selected based on manual empirical testing, and was thus exhaustively searched:
\begin{itemize}
    \item Dropout $\in \{0, 0.25, 0.5\}$
    \item L2 regularization $\lambda \in \{0, 0.0003, 0.001, 0.003, 0.01\}$
    \item Number of dense layers $\in \{2, 3, 4\}$
    \item Optimizer = Adam \cite{adam}
    \item Loss = mean squared error (found empirically to outperform binary cross-entropy)
    \item Activation = tanh (found empirically to outperform ReLU)
    \item Batch size = 32
\end{itemize}

@@

Because all data in the PICO-60 experiment is labeled, unlabeled training data sets are randomly sampled from the 496 training examples remaining (after the validation set is selected). Labeled set sizes of 128 and 256 were tested; the remainder of the 496 examples would be used as unlabeled data.

@@

It is a transformation of the hyperbolic tangent that flattens the central range and comparatively exaggerates the asymptote on either side, as visualized in Figure \ref{grav_diff}. The equation above can be described in the following steps:

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.85\textwidth]{grav_diff}
    \caption{\label{grav_diff} A visualization of $\mathrm{GravDiff}(p, \psi, g)$ with respect to $p$ and $\psi$.}
\end{figure}

@@

Table \ref{results_table} compares performance on test data of the best-performing architectures for each algorithm and training set size. These architectures were selected based on accuracy on validation data.

\begin{minipage}{\textwidth}
    \begin{center}
        \begin{tabular}{|l|l|l|l|l|l|}
            \hline
            Algorithm & Initial $\varsigma$ Size & Accuracy & Precision & Recall \\
            \hline
            Original PICO-60 ML & 278 & 80.7\% & 76.9\% & 100.0\% \\
            \hline
            Acoustic Parameter & N/A & 100\% & 100\% & 100\% \\
            \hline
            Supervised & 128 & 96.79\% & 95.85\% & 99.09\% \\
            \hline
            Supervised & 256 & 96.97\% & 95.99\% & 99.59\% \\
            \hline
            Pseudo-Labels & 128 & 94.89\% & 93.31\% & 99.36\% \\
            \hline
            Pseudo-Labels & 256 & 97.01\% & 95.30\% & 100.00\% \\
            \hline
            Gravitational Differentiation & 128 & 94.78\% & 96.68\% & 99.28\% \\
            \hline
            Gravitational Differentiation & 256 & 98.29\% & 98.75\% & 99.44\% \\
            \hline
        \end{tabular}
        \captionof{table}{Summary of the most successful semi-supervised learning models.} \label{results_table}
    \end{center}
\end{minipage}

It can be seen in Table \ref{results_table} that with 128 labeled examples and 368 unlabeled examples, supervised learning outperforms both pseudo-labels and gravitational differentiation. This is likely due to the semi-supervised learning algorithms reinforcing non-representative biases present in the small set of data.

However, when 256 labeled examples and 240 unlabeled examples are used, it is observed that pseudo-labels performs almost identically to supervised learning, and that gravitational differentiation outperforms both existing algorithms. This indicates that, when the right balance between labeled and unlabeled examples is used, the confidence of predictions on unlabeled examples as used in gravitational differentiation is a meaningful metric and can be used to further optimize a neural network alongside a conventional loss function.

The test set predictions of gravitational differentiation trained on 256 labeled examples can be seen in Figure \ref{grav_test_hist}, with Acoustic Parameter on the X axis for comparison. The decision boundaries of AP and the neural network are represented by black lines.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{grav_test_hist}
    \caption{\label{grav_test_hist} Test set predictions of a gravitational differentiation model, compared against AP.}
\end{figure}

@@

This study has demonstrated that it is possible, using semi-supervised learning, to automatically optimize a discriminator with performance comparable to one developed and tuned by humans like the AP technique. This conclusion has been achieved through two rounds of optimization: the first to explore and evaluate various network structures and input formats using supervised learning, and the second to incorporate the most successful configuration into semi-supervised learning systems with the goal of producing a high-accuracy discriminator to complement AP.
